---
title: "EM Algorithm on Missing Data"
output: html_notebook
---
## Dataset Generation 
We will generate 1000 samples for the Systolic Blood Pressures each of size 50 for the Month of January and February. 

```{r}
library(MASS)

# Set parameters for the bivariate normal distribution

set.seed(0)
mu <- c(125, 125)  # Mean vector

sd_x <- 25
sd_y <- 25
cov <- 0.6*sd_x*sd_y

sigma <- matrix(c(625, cov, cov, 625), nrow = 2)  # Covariance matrix


n <- 50 # Number of individuals
num_samples <- 1000 # Number of samples

# Storage for the samples
all_samples <- list()

# Generate random samples iteratively from a bivariate normal distribution. 
for (i in 1:num_samples) {
  samples <- mvrnorm(n = n, mu = mu, Sigma = sigma)
  all_samples[[i]] <- samples
}

# Check the first few observations of the first sample
print(head(all_samples[[1]]))

```


'all_X' contains the data for the month of January. 'all_Y' is the data for the month of February. 

```{r}
all_X <- list()
all_Y <- list()

for (i in 1:num_samples) {
  X_sample <- all_samples[[i]][, 1]
  Y_sample <- all_samples[[i]][, 2]
  all_X[[i]] <- X_sample
  all_Y[[i]] <- Y_sample
}

# Check the first few observations of X and Y for the first sample
print("X for the first sample:")
print(head(all_X[[1]]))
print("Y for the first sample:")
print(head(all_Y[[1]]))
```

Calculating the Mean of the complete data. 

```{r}
# Take the Mean of all the samples 
mux <- rep(0, num_samples)
muy <- rep(0, num_samples)

for (i in 1:num_samples){
  
  mux[i] <- mean(all_X[[i]])
  muy[i] <- mean(all_Y[[i]])
  
}
mu_X = mean(mux)
mu_Y = mean(muy)

cat("Mean for the month of January is ", mu_X)
cat("\nMean for the month of February is ", mu_Y)

```

# Creating Missing Data
Given the complete data we will now impose the conditions of Missing Completely At Random (MCAR), Missing At Random (MAR) and Missing Not At Random (MNAR) to generate missing data. Here, we create missing data only in the data of February month. 

### Missing Completely At Random (MCAR)

```{r}
set.seed(0)
# We implement the code to get MCAR Data for the month of February. 

# Probability of missingness
p_missing <- 0.8

Y1 <- list()

for(i in 1:num_samples){
# Generate MCAR dataset Y1
Y1[[i]] <- ifelse(runif(length(all_Y[[i]])) < p_missing, NA, all_Y[[i]])
}

```


### Missing At Random (MAR)
```{r}

Y2 <- list()
threshold <- 140
for(i in 1:num_samples){
  Y2[[i]] <- ifelse(all_X[[i]] < threshold, NA, all_Y[[i]])
}
```


### Missing Not At Random (MNAR)
```{r}
Y3 <- list()

threshold <- 140
for(i in 1:num_samples){
  Y3[[i]] <- ifelse(all_Y[[i]] < threshold, NA, all_Y[[i]])
  
}
```



# Data Imputation
Given the missing data, we will the question will arises, as to which Data Imputation method should be used in each of the three cases of Missing Data.


### Expectation Maximization (EM) Algorithm
The EM Algorithm aims at finding the estimate of the parameter (in the bivariate normal case it is $\mu$ and $\Sigma$) that maximizes the log-likelihood of the observed data. 

As the name suggests, the process consists of two steps:

1. Expectation Step - This computes the expected value of the log-likelihood function given the observed data. i.e. it calculates the conditional expectation using the parameter estimate of the current iteration. 

2. Maximization Step - This step computes the maximum likelihood estimate over the expectation computed in step-1. 

This process is repeated iteratively till it achieves convergence. 

```{r}

# Expectation Step of the algorithm
Estep <- function(all_X, all_Y, mu, sigma){
  num_samples <- length(all_X)
  
  # Initialize sums
  sum_E_x <- sum_E_y <- sum_E_x_x <- sum_E_y_y <- sum_E_x_y <- 0
  
  for (i in 1:num_samples) {
    X <- all_X[[i]]
    Y <- all_Y[[i]]
    
    missing_idx <- which(is.na(Y))
    observed_idx <- which(!is.na(Y))
    
    
    # Computed using the likelihood estimation method for bivariate conditional gaussian distribution. 
    beta_21.1 = sigma[1,2] / sigma[1,1]
    beta_20.1 = mu[2] - beta_21.1 * mu[1]
    sigma_22.1 = sigma[2,2] - sigma[1,2]^2 / sigma[1,1]
    
    # Impute the Data
    E_y = rep(0, length(Y))
    E_y[missing_idx] = rep(beta_20.1, length(missing_idx)) + beta_21.1 * X[missing_idx]
    E_y[observed_idx] = Y[observed_idx]
    
    E_x = X
    
    E_y_y = rep(0, length(Y))
    E_y_y[missing_idx] = rep(sigma_22.1, length(missing_idx)) + E_y[missing_idx]^2
    E_y_y[observed_idx] = E_y[observed_idx]^2
    
    E_x_x = X^2
    
    E_x_y = rep(0, length(Y))
    E_x_y = E_y * E_x 
    
    # Update sums
    sum_E_x = sum_E_x + mean(E_x)
    sum_E_y = sum_E_y + mean(E_y)
    sum_E_x_x = sum_E_x_x + mean(E_x_x)
    sum_E_y_y = sum_E_y_y + mean(E_y_y)
    sum_E_x_y = sum_E_x_y + mean(E_x_y)
  }
  
  
  return(structure(list(s1 = sum_E_x, s2 = sum_E_y, s11 = sum_E_x_x, s22 = sum_E_y_y, s12 = sum_E_x_y)))
}


# Maximization Step of the Algorithm
Mstep <- function(X, s1, s2, s11, s22, s12){
  n <- length(X)
  
  muX <- s1 / n
  muY <- s2 / n
  
  sigma1 <- s11 / n - muX^2
  sigma2 <- s22 / n - muY^2
  sigma12 <- s12 / n - muX * muY
  
  mu <- c(muX, muY)
  
  cov_mat <- matrix(c(sigma1, sigma12, sigma12, sigma2), nrow = 2)
  
  return(structure(list(mu = mu, cov_mat = cov_mat)))
}


# Initializing the mean vector and the covariance matrix for the incomplete data. 

initEM <- function(all_X, all_Y){
  num_samples <- length(all_X)
  
  # Initialize sums
  sum_X <- sum_Y <- sum_X_X <- sum_Y_Y <- sum_X_Y <- 0
  
  for (i in 1:num_samples) {
    X <- all_X[[i]]
    Y <- all_Y[[i]]
    
    missing_idx <- which(is.na(Y))
    observed_idx <- which(!is.na(Y))
    
    sum_X <- sum_X + mean(X)
    #print(sum_X)
    sum_Y <- sum_Y + mean(Y[observed_idx])
    sum_X_X <- sum_X_X + mean(X^2)
    sum_Y_Y <- sum_Y_Y + mean(Y[observed_idx]^2)
    sum_X_Y <- sum_X_Y + mean(X[observed_idx] * Y[observed_idx])
  }
  
  muX <- sum_X / num_samples
  muY <- sum_Y / num_samples
  
  
  varx <- sum_X_X / num_samples - muX^2
  vary <- sum_Y_Y / num_samples - muY^2
  
  cov_xy <- sum_X_Y / num_samples - muX * muY
  
  mu <- c(muX, muY)
  
  cov_mat <- matrix(c(varx, cov_xy, cov_xy, vary), nrow = 2)
  
  return(structure(list(mu = mu, cov_mat = cov_mat)))
}


```



```{r}
# Function to run the EM Algorithm for 150 iterations. 

runEM <- function(X, Y){
  
  init = initEM(X, Y)
  
  mu_hat = init$mu
  cov_matrix = init$cov_mat
  error_mu=rep(0,150)
  for(i in 1:150){
    error_mu[i]=sqrt(sum((mu_hat-mu)^2))
    # E step
    E=Estep(X, Y, mu_hat, cov_matrix)
    s1=E$s1
    s11=E$s11
    s2=E$s2
    s22=E$s22
    s12=E$s12
    
    # M Step 
    M=Mstep(X, s1, s2, s11, s22, s12)
    mu_hat=M$mu
    #print(mu_hat)
    cov_matrix = M$cov_mat
  }
  
  
  #corr = cov_matrix[1,2]/(sqrt(cov_matrix[1,1]) * sqrt(cov_matrix[2,2]))
  
  #print(mu_hat)
  #plot(error_mu)
  
  
  return(structure(list(mu_hat = mu_hat, cov_matrix = cov_matrix, error_mu = error_mu)))
}

```


We run the EM Algorithm for the MCAR Data
```{r}
res1 = runEM(all_X, Y1)
mu1 = res1$mu_hat
cov1 = res1$cov_matrix
corr1 = cov1[1,2]/(sqrt(cov1[1,1]) * sqrt(cov1[2,2]))


cat("The Estimated Mean in the MCAR Data is ", mu1[2])
cat("\nThe Estimated Variance in the MCAR Data is ", sqrt(cov1[2,2]))
cat("\nThe Estimated Correlation in the MCAR Data is  ", corr1)

```

We run the EM Algorithm for the MAR Data
```{r}
res2 = runEM(all_X, Y2)
mu2 = res2$mu_hat
cov2 = res2$cov_matrix
corr2 = cov2[1,2]/(sqrt(cov2[1,1]) * sqrt(cov2[2,2]))


cat("The Estimated Mean in the MCAR Data is ", mu2[2])
cat("\nThe Estimated Variance in the MCAR Data is ", sqrt(cov2[2,2]))
cat("\nThe Estimated Correlation in the MCAR Data is  ", corr2)
```


We run the EM Algorithm for the MNAR Data. 
```{r}
res3 = runEM(all_X, Y3)
mu3 = res3$mu_hat
cov3 = res3$cov_matrix
corr3 = cov3[1,2]/(sqrt(cov3[1,1]) * sqrt(cov3[2,2]))


cat("The Estimated Mean in the MCAR Data is ", mu3[2])
cat("\nThe Estimated Variance in the MCAR Data is ", sqrt(cov3[2,2]))
cat("\nThe Estimated Correlation in the MCAR Data is  ", corr3)

```

